The not so widening achievement gap: an interntational comparison of the evolution of the SES achievement gap from 2000-2015

### 2017-09-08 paper/paper.Rnw
Revised all the vocabulary/punctuation - further analysis, graphs and text left

Aside from SES, the other most relevant variables are test scores for Mathematics and Literacy. PISA does not provide a single test result for each respondent. Instead, it provides a \emph{series} of 'plausible values' that the child could actually score. As explained in the PISA manual \citep{pisa2012_technical}, these are imputed values that resemble individual test scores and have approximately the same distribution as the latent trait being measured (the true distribution of the possible scores a student can achieve).

A more intuitivity is explanation is this: suppose we have \(\mu_i\), the average student test score in Mathematics for student \(i\). Instead of estimating \(\mu_i\) alone, plausible values estimate a distribution of possible \(\mu\text{'s}\) for student \(i\), together with the likelihood of each \(\mu_i\) based on the respondents answers on the test. This is defined as the posterior distributions of \(\mu\text{'s}\) for student \(i\). The reason why we use this procedure is because estimating a single estimate \(\mu_i\) is plagued with measurement error, among other types of bias \citep[see][]{wu2005}. The number of plausible values for PISA waves are usually five (although ten for PISA 2015) random draws from this distribution. In practice each student has 5 scores for each test, that resemble their distribution. Those values are continuous, ranging from 0 to 500, with a mean of 250.

@@

The aim of this paper is to identify and disaggregate country trends in the achievement gap for several countries. To represent the SES gap, most of the literature on achievement gaps has concentrated on indicators such as parental education, parental occupational status, income achievement gaps and actual SES achievement gaps \citep{fryer2004, hanushek_woesmann_tracking, saw2016, bradbury2015, byun2010}. The actual calculation of the achievement gap varies substantially and different strategies have been implemented. For example, \citet{micklewright} calculates the difference in achievement by crudely subtracting the gap between the 95th and 5th percentile of the mathematics distribution. Although in principle you should be able to capture some type of SES effect like this, theoretically, it should be much more accurate to difference out the mean score of, for example, parental education or some other SES proxy. \citet{saw2016}, for instance, used parental education as a proxy of SES, whereas \citet{byun2010} use a similar SES index as ours, but created by them.

\citet{reardon_portilla}, \citet{anna2016} and \citet{anna2016_global} used a different method developed by \citet{reardon2011} which we partially adopt in this paper. SES achievement gaps are measured as the difference in standardized achievement between the 90th and 10th percentiles of the chosen SES variable. The rule of thumb to choose the 90th, 50th and 10th percentile is arbitrary, as others have used, for example, the 5th, 50th and 95th \citep{micklewright}. We use the conventional 90/10 cutoffs in the literature following the standard set by \citet{reardon2011}.

@@

for each wave, where \begin{math}Y_i\end{math} is the median student test score and \begin{math}AGE_i\end{math} is their age measured in months (following the same strategy as \citet{reardon2011}) weighted by the student sample weights.

@@

In order to correct gap estimates for measurement error, I adjust student test scores(\begin{math} \hat{\gamma_i} \end{math}) by

@@

where \begin{math}r\end{math} is the reliability score of the wave \footnote{Other proceedures multiply each country by their own reliability measure for each year-subject pair \citep{anna2016_global}. The realibility estimates are calculated using Item Response Theory (IRT) analogues of traditional estimates of Person separation reliability such as internal consistency. Unfortunately, PISA 2000 did not provide any reliability measure separately for each country and PISA 2015 has to yet release their own. At the moment of writing this paper, they were unavailable. For this reason we implement the analysis following the original work of \citet{reardon2011}}. Each PISA survey provides a reliability indicator which we use accordingly. This yields estimates of the true gaps, and eliminates any bias in the trend that may arise from differential reliability of the tests.

@@

We see that in some countries like Finland and Sweden the average low SES scores is actually above the average score of 0, suggesting their very equal countries. However, most countries, like Spain, Poland, Hungary and Germany, their average low SES scores are much lower than the country average, which is 0. We also see some countries with major changes from year 2000 to 2015, with, for example, Australia decreasing the low SES gap from 0.22 standard deviations to -0.15, well below the country average. In the next section we take a look at this in a more detailed fashion, considering all years in between.

@@

We start by look at the achievement for some countries in figure 2. We plot only mathematics for each country and also a quadratic trend spline for \emph{both mathematics and literacy} pooled. Some countries have increased their achievement very strongly. For example, France, Austria and surprisingly Sweden have very steep slopes. France increasde the gap by roughly \Sexpr{diff_data$math[["France"]]} standard deviations, Austria by \Sexpr{diff_data$math[["Austria"]]} and Sweden by \Sexpr{diff_data$math[["Sweden"]]}. This pattern happens similarly for literacy. For example, France has an increase of \Sexpr{diff_data$read[["France"]]}. For such a short period of time, the magnitud of these increases are reasonably big.

@@

Other countries have reasonable increases such as Finland and Hungary, with increases of nearly \Sexpr{diff_data$math[["Finland"]]} and \Sexpr{diff_data$math[["Hungary"]]}, respectively. Aside from these countries, there are other cases that have no change at all, specifically, Canada, Netherlands and Spain. Canada excels here not only because the gap has been stable over time, but because it has the smallest gap of all countries presented here. It's nearly 0.5 standard deviations in 2000 and it increased only by \Sexpr{diff_data$math[["Canada"]]} in 2015. We can't really say Italy has had a significant since it has a sort of wiggly pattern that looks like it cancels itself out.

On the other hand, we do have some countries which show a decrease in the SES achievement gap. United Kingdom decreased by about \Sexpr{diff_data$math[["United Kingdom"]]}, Poland by \Sexpr{diff_data$math[["Poland"]]}, and Denmark by \Sexpr{diff_data$math[["Denmark"]]}. However, the most notable cases are the United States and Germany. These two countries show a very high level of dispersion in the year 2000 with SES gaps of over 2 standard deviations. But in the 15-year time trend both countries reduced the gaps by \Sexpr{diff_data$math[["United States"]]} and \Sexpr{diff_data$math[["Germany"]]} respectively. Their distinctively big gaps in 2000 also show up in the work of \citet{micklewright}. This corroborates the findings of \citet{reardon_portilla}, which found a decreasing gap for kindergarteners. This decline is evident also for 15 year olds, suggesting it might be more of an institutional change rather than a specific grade-level policy.

Analyzing this graph we might naively conclude that these trends are not very steep and they should not be very important to consider. However, remember that the Y axis is measured in standard deviations. Small changes are actually very huge in practical terms. Take the case of Sweden. The slope doesn't look that steep but in reality it increased the gap from 1 standard deviation in 2000 to around 1.5 standard deviation in 2015. The actual difference is of about \Sexpr{diff_data$math[["Sweden"]]}. Now that we know that, the trends of Poland, United States, France and Germany are particularly shocking.

@@

us_math_2000_upper <-
  filter(complete_data_topbottom, wave == 2000, country == "United States", type_test == "math") %>%
  .[["upper"]]
@
We find that the initial gap for the U.S in 2000 is \Sexpr{round(us_math_2000, 2)} varying between \Sexpr{round(us_math_2000_lower, 2)} and \Sexpr{round(us_math_2000_upper, 2)}. This gap is much higher than found in previous studies, but again, we don't know much previous estimates vary and nobody has really studied the 15-year old achievement gap precisely. All country-level studies pool many studies together and control for age, removing the age-specific gap effect.

@@

Something reassuring is that mathematics and reading (not presented) follow basically the same trend across all countries \footnote{Reading had an even bigger decrease for similar countries, and a much smaller increase. The previos two graphs for reading are available upon request}. This means that the result is not an artifact of chance. However, it's important to disentangle where is this gap originating from. Is this because the top are improving why the bottom decreases? Or is it that the bottom is catching up? Up next we plot the same graph but show the divergent patterns between high/low SES origins.

@@

Figure 4 shows the evolution for both Low and High SES groups separetly. The middle line is a summary line showing whether one of the groups is growing/shrinking faster than the other. We find plenty of variation in these patterns. For example, in the United States the top seems to be equalizing much stronger than than the bottom is catching up. The U.K seems to be following the same path as the United States as well. On the other hand, in Poland the gap seems to be closing because the bottom SES group is catching up much faster. The Netherlands shows a similar pattern as the U.S, U.K patterns but upon closer inspection the explanation is very different. We see that the slope is negative (like in the previous two countries) but that's because the low SES group is decreasing at a greater rate than the high SES is decreasing.

@@

We see that there are more countries where High SES is increasing inequality faster than low SES is catching up (red panel) but there's also a fair share where low SES is catching up faster, such as the United States and Germany. We see a negative relationship where the faster the low SES catch up the slower the High SES decrease. This is very interesting because it suggests that there's a trend for countries to have the bottom SES groups are climbing upwards, a trait of meritocracy. Equally important, we see very few countries where low SES are decreasing (so going even further down) and High SES is also decresing (bottom left panel). On the other hand, we do see more ocuntries where the both groups are increasing (top right panel). We plot all remaining countries in the back, together with a trendline and we see that the relation holds for all PISA countries.

@@

Given that we have the data for the average performance for each country and the achievement gap, we can also investigate whether there's a tradeoff between the average country performance and the widening of the achievement gap and if it stands over time. One particular topic that has been researched in the past is whether improving the overall performance of a country compromises the equality of the educational system. For example, in order to improve the average score in, let's say, mathematics, tracking can improve the average score by making children in the higher trackers perform better, raising the overall performance. This, however, can come at a price. If tracking actually benefits the top but doesn't increase the performance of the bottom tracks then we see a net increase of inequality accompanied with an increase in the overall performance. We plot this relationship in figure 8.

\subsection{Performance-equality tradeoff}
<<graphing_achievement_disparity, cache = T>>=

@@

For the past 15 years we see a very strong correlation suggesting that countries that perform very well in PISA have actually very low achievement gaps (or the other way around) \footnote{We graph all countries-years pooled as well and the relationship is identical}. The correlations are between \Sexpr{round(max(correlation), 2)} and \Sexpr{round(min(correlation), 2)}. This confirms the theoretical and empirical frameworks suggested by \citet{werfhorst_mijs}. However, what they didn't really go in depth was whether this corelation is strongest for some countries over time. With the data available we can look at the evolution separately for many countries. We plot it up next in figure 9.

@@

Looking at figure 9, we see a strong correspondence between the two variables. In countries like France we see quite a functional form between both inequality and performance. As average performance plummets, the achievement gap is increasing. This pattern is also seen in Austria, Australia and Germany. The other obvious pattern is the opposite trend. We see countries like Poland, Finland, Canada and less pronounced, Italy, where this pattern is evident. The surprising case here is the United States. We see that as the average performance has been decreasing, the average achievement gap has been decreasing as well. In fact, among the few countries we should find something that resembles this pattern should be the U.S because it doesn't have a strict institutionalized tracking system. However, the explanation behind this pattern is very speculative.

@@

\section{The role of curricular tracking in explaining inequalities}

@@

The reason why countries differ in their evolution of inequality is still unknown. There is ample evidence showing the inequality between countries, to a certain extent, can be explained by the degree of tracking (Hanushek and Woessman). But we are also interested in what explains the between country evolution of inequality, that is, the explanation as to why in certain countries it is increasing more than others over time.

This is clearly a difficult task given than, at least the country-level, not many features of the educational system change in a 15-year time span, resulting in a practically invariant variable from which to explain the level of inequality (without any variance, we can't explain anything.). However, there are other strategies that can at least partially help us to understand the phenomena, and these are precisely the ones we follow.

Following the literature on the topic, we suspect that this evolution is in part very tightly related to the degree of stratification of the educational system. In a very simple frame-work, Hanushek and Woessman found that tracking indeed exarcebates inequality in the life course of children. Following the same strategy, we believe that tracking is also allowing the same gap in 15 year olds to grow over time, acompannied by other things. Yet, instead of concetrating only on the age of selection (just as they do), we explore a more fine grained definition of tracking, which is possible through the work of Thijs and Van der Werfhorst(). Excluding the age of first selection into tracks, we also use the the number of tracks in the country, the percentage of the entire curriculum that is tracked and a vocational index \footnote{This index is a factor loading from a principal factor analysis of the percentage of students in upper secondary vocational education (taken from two sources, to reduce measurement error)}.

@@

In the final model, we regress the inequality of a country on a dummy variable where 1 equals only 1 track vs more than 1 track, whether the age of first selection is 15 or more vs below 15, a dummy stating whether the percentage of tracked curriculum is above 1, and the standardized vocational index. We are reluctant to include more variable in the model, first to keep it more parsimonious, and secondly to prevent overfitting due to our sample size. We present the results in the table below.

@@

Given that the classical p values are inexistent in bayesian modelling, I present uncertainty intervals next to each coefficient. The first model only includes the 1 track dummy. This variable shows that countries that have one track, have on average, half a standard deviation less inequality than other countries. This covariate alone explains about 23\% of the variance of the evolution of the 90/10 achievement gap. The second model now includes the dummy of whether the country has an age selection of 15 or more and we can see that countries with late selection into tracking have about -.38 less standard deviations in inequality. Also note that the number of tracks dummy was also reduced, suggesting these two variables are explaining similar things. This is expected given that late tracking is usually associated with fewer tracks, such as in the scandinavian countries. The third model now includes whether the country has any degree of tracking, meaning whether the country has above 0\% of the curriculum tracked. The coding of the variable comes from visual inspection of the distribution, where some countries, such as the United States and Canada, have untracked curriculums. We see now relationship at all, but after including the vocational index, the model is completely reversed with countries with tracked curriculum having about 0.6 more standard deviations in inequality that countries with no tracking.

We also note that vocational enrollment is associated with less inequality. Not surprisingly, the age of selection coefficient, also increased due to the correlation between age of selection and vocational enrollment. Countries with high vocational enrollment are also those with early age selection, such as Germany and Austria. These two variables should be inspected further, perhaps with an interaction, given than both coefficients are sizable and very correlated. This final model explains over one third of the variance in the evolution of the 90/10 SES gap, with an R square of 40\%. The uncertainty of each of the covariates varies a lot with the percentage of curriculum tracked dummy ranging all the way from 0.9 to 0.3. Nonetheless, we see that the other coefficients are very strong and have reasonable intervals. It is very important to consider the uncertainty of our estimates, to give confidence in our model fit. So far, this model seems to fit the data quite reasonably. \footnote{These variables are very correlated, but no collinearity is present with a maximum VIF of 2.2.}

All of these countries have a different tracking structure, with some setups being more egalitarian than others. Using the previous model we make a simulation and predict the level of the 90/10 achievement gap for all possible tracking setups for all countries. Below we plot all these combinations. The x axis has the number of tracks that a country has, the y axis has the predicted 90/10 achievement gap and below each country name we have whether the country has an age selection of 15 or more or below. In addition, to confirm how accurate is the model, we also plot the actual level of inequality in the actual tracking setup of the country. Let's take Germany for example.

@@

The red point for Germany is in the '\textgreater 1 track' and '\textless 15' setup, meaning that Germany has a tracking system that has more than 1 track and an age of selection of below 15. We see that the model predicts the level of inequality for that set up quite accurately, relative to the actual level (red point). But if the setup would be in the ideal '1 track' and age of selection above or equal to 15, the inequality would be much lower. Generally speaking, this applies to virtually every country in the plot. Virtually for all countries which are not in this ideal tracking, a switch to the ideal would reflect a reduction in the achievement gap.

For those countries which are not in the 'ideal tracking' setup, we compute below the percentage change from switching to the ideal tracking. The results are in the table below.

@@

The biggest reductions are for both Hungary and Germany, with 46\% and 36\% respectively. France, which had a particularly steep increase in the last 15 years, would see a reduction of the 90/10 SES gap of 31\% and Austria, with a very similar tracking structure as the German one, would see their achievement gap reduced by 20\%. The very suprising result is the Dutch case, for which a switch from the 'worst' tracking setup to the 'best' case would result in an increase of the gap of 20\%.

This result could mean two things. First and foremost, the Dutch educational system has built other mechanisms that, despite the tracking system, help children from social origin escalate socially. This explanation is not very speculative given than in most mobility research, the Netherlands stands out as a very equal country relative to other countries with a similary tracking structure, like Germany or Hungary (cite Gosta study and something from Breen and Goldthorpe). Another explanation, and something which needs to assessed, is whether our model is predicting accurately. If we look at the plot from above, we see that for some countries, the model is a bit off from the actual level of inequality. Moreover, although we see that if all countries switched to the 'ideal' tracking setup we would see a vast reduction in inequality, we also see that each of these predictions is followed by a wide uncertainty interval. Considering that the results are very consistent for every country, and in most countries the prediction matches the actual value, this could simply be a consequence of low sample size. Whether these results stand the test of a different empirical strategy or a higher sample size, we leave up to future research.

A more summarized empirical strategy is to actually standardize all the tracking indicator into one index, something that Bol and Van der Werfhorst() already do quite nicely. Using this standardized tracking variable and the same bayesian multievel model, we now model the tracking and vocational index as before, but add an interaction between two, given that in the previous model we recorded a strong shift in the model when tracking-related variables and the vocational model we both included in the same model. Below are the results for this model.

@@

We confirm the results from the previous models, where more tracking is associated with an achievement gap of about 0.4 standard deviations wider and more vocational enrollment is associated with a reduction of about 0.2 standard deviations, half the slope of tracking. Meaning that the tracking effect might be offsetting the effect of the vocational index given that it's half the size of the slope. Both these coefficients are actually significantly different from each other, so it is possible. Before we explore this visually, let's note that the interaction between the two is very reliable and significant, considering that small sample size. Below we graph the interaction for different quantiles of the vocational index.

>>>>

Aside from SES, the other most relevant variables are test scores for mathematics and literacy. PISA does not provide a single test result for each respondent. Instead, it provides a \emph{series} of 'plausible values' that the child could actually score. As explained in the PISA manual \citep{pisa2012_technical}, these are imputed values that resemble individual test scores and have approximately the same distribution as the latent trait being measured (the true distribution of the possible scores a student can achieve).

A more intuitivity is explanation is this: suppose we have \(\mu_i\), the average student test score in mathematics for student \(i\). Instead of estimating \(\mu_i\) alone, plausible values estimate a distribution of possible \(\mu\text{'s}\) for student \(i\), together with the likelihood of each \(\mu_i\) based on the respondents answers on the test. This is defined as the posterior distributions of \(\mu\text{'s}\) for student \(i\). The reason why we use this procedure is because estimating a single estimate \(\mu_i\) is plagued with measurement error, among other types of bias \citep[see][]{wu2005}. The number of plausible values for PISA waves are usually five (although ten for PISA 2015) random draws from this distribution. In practice each student has 5 scores for each test, that resemble their distribution. Those values are continuous, ranging from 0 to 500, with a mean of 250.

@@

The aim of this paper is to identify, disaggregate and explain country trends in the achievement gap for several countries. To represent the SES gap, most of the literature on achievement gaps has concentrated on indicators such as parental education, parental occupational status, income achievement gaps and actual SES achievement gaps \citep{fryer2004, hanushek_woesmann_tracking, saw2016, bradbury2015, byun2010}. The actual calculation of the achievement gap varies substantially and different strategies have been implemented. For example, \citet{micklewright} calculates the difference in achievement by crudely subtracting the gap between the 95th and 5th percentile of the mathematics distribution. Although in principle you should be able to capture some type of SES effect like this, theoretically, it should be much more accurate to difference out the mean score of, for example, parental education or some other SES proxy. \citet{saw2016}, for instance, used parental education as a proxy of SES, whereas \citet{byun2010} use a similar SES index as ours, but created by them.

\citet{reardon_portilla}, \citet{anna2016} and \citet{anna2016_global} used a different method developed by \citet{reardon2011} which we partially adopt in this paper. SES achievement gaps are measured as the difference in standardized achievement between the 90th and 10th percentiles of the chosen SES variable. The rule of thumb to choose the 90th, 50th and 10th percentile is arbitrary, as others have used, for example, the 95th, 50th and 5th \citep{micklewright}. We use the conventional 90/10 cutoffs in the literature following the standard set by \citet{reardon2011}.

@@

for each wave, where \begin{math}Y_i\end{math} is the median student test score and \begin{math}AGE_i\end{math} is their age measured in months (following the same strategy as \citet{reardon2011} \footnote{This does not mess up our analysis by masking age-specific gaps because all students in the sample are 15 year olds. Control for aging is simply to control for monthly differences in ages.}) weighted by the student sample weights.

@@

In order to correct the gap estimates for measurement error, we adjust student test scores(\begin{math} \hat{\gamma_i} \end{math}) by

@@

where \begin{math}r\end{math} is the reliability score of the wave \footnote{Other proceedures multiply each country by their own reliability measure for each year-subject pair \citep{anna2016_global}. The realibility estimates are calculated using Item Response Theory (IRT) analogues of traditional estimates of person separation reliability such as internal consistency. Unfortunately, PISA 2000 did not provide any reliability measure separately for each country and PISA 2015 has to yet release their own. At the moment of writing this paper, they were unavailable. For this reason we implement the analysis following the original work of \citet{reardon2011}}. Each PISA survey provides a reliability indicator which we use accordingly. This yields estimates of the true gaps, and eliminates any bias in the trend that may arise from differential reliability of the tests.

@@

We see that in some countries like Finland and Sweden the average low SES scores is actually above the average score of 0, suggesting they're very equal countries. However, most countries, like Spain, Poland, Hungary and Germany, their average low SES scores are much lower than the country average, which is 0. We also see some countries with major changes from year 2000 to 2015, with, for example, Australia decreasing the low SES gap from 0.22 standard deviations to -0.15, well below the country average. In the next section we take a look at this in a more detailed fashion, considering all years in between.

@@

We start by look at the achievement for some countries in figure 2. We plot only mathematics for each country and also a quadratic trend spline for \emph{both mathematics and literacy} pooled. Some countries have increased their achievement very strongly. For example, France, Austria and surprisingly Sweden have very steep slopes. France increased the gap by roughly \Sexpr{diff_data$math[["France"]]} standard deviations, Austria by \Sexpr{diff_data$math[["Austria"]]} and Sweden by \Sexpr{diff_data$math[["Sweden"]]}. This pattern happens similarly for literacy. For example, France has an increase of \Sexpr{diff_data$read[["France"]]}. For such a short period of time, the magnitud of these increases are reasonably big.

@@

Other countries have reasonable increases such as Finland and Hungary, with increases of nearly \Sexpr{diff_data$math[["Finland"]]} and \Sexpr{diff_data$math[["Hungary"]]}, respectively. Aside from these countries, there are other cases that have no change at all, specifically, Canada, Netherlands and Spain. Canada excels here not only because the gap has been stable over time, but because it has the smallest gap of all countries presented here. It's nearly 0.5 standard deviations in 2000 and it increased only by \Sexpr{diff_data$math[["Canada"]]} in 2015. We can't really say Italy has had a significant increase since it has a sort of wiggly pattern that looks like it cancels itself out.

On the other hand, we do have some countries which show a decrease in the SES achievement gap. Poland decreased by about \Sexpr{diff_data$math[["Poland"]]} and Denmark by \Sexpr{diff_data$math[["Denmark"]]}. However, the most notable cases are the United States and Germany. These two countries show a very high level of dispersion in the year 2000 with SES gaps of over 2 standard deviations. But in the 15-year time trend both countries reduced the gaps by \Sexpr{diff_data$math[["United States"]]} and \Sexpr{diff_data$math[["Germany"]]} respectively. Their distinctively big gaps in 2000 also show up in the work of \citet{micklewright}. This corroborates the findings of \citet{reardon_portilla}, which found a decreasing gap for kindergarteners. This decline is evident also for 15 year olds, suggesting it might be more of an institutional change rather than a specific grade-level policy. Do note that the trendline is for both mathematics and literacy. If it were only for mathematics, then the trendline would probably be even more extreme. Despite this, pooling both subjects gives even more strength to the analysis given that the results stand for both subjects together \footnote{Same graph is availabe for reading upon request, where we find pretty much the same results if not even more steep/downward slopes.}.

Analyzing this graph we might naively conclude that these trends are not very steep and they should not be very important to consider. However, remember that the Y axis is measured in standard deviations. Small changes are actually very huge in practical terms. Take the case of Sweden. The slope doesn't look that steep but in reality it increased the gap from 1 standard deviation in 2000 to around 1.5 standard deviation in 2015. Now that we know that, the trends of Poland, United States, France and Germany are particularly shocking.

@@

us_math_2000_upper <-
  filter(complete_data_topbottom, wave == 2000, country == "United States", type_test == "math") %>%
  .[["upper"]]
@

We find that the initial gap for the U.S in 2000 is \Sexpr{round(us_math_2000, 2)} varying between \Sexpr{round(us_math_2000_lower, 2)} and \Sexpr{round(us_math_2000_upper, 2)}. This gap is much higher than found in previous studies, but again, we don't know how much previous estimates vary and nobody has really studied the 15-year old achievement gap precisely. All country-level studies pool many studies together and control for age, removing the age-specific gap effect.

@@

Something reassuring is that mathematics and reading (not presented) follow basically the same trend across all countries \footnote{Reading had an even bigger decrease for similar countries, and a much smaller increase. The previos two graphs for reading are available upon request}. This means that the result is not an artifact of chance. However, it's important to disentangle where is this gap originating from. Is this because the top are improving while the bottom decreases? Or is it that the bottom is catching up? Up next we plot the same graph but show the divergent patterns between high/low SES origins.

@@

Figure 4 shows the evolution for both low and high SES groups separately. The middle line is a summary line showing whether one of the groups is growing/shrinking faster than the other. We find plenty of variation in these patterns. For example, in the United States the top seems to be equalizing much stronger than than the bottom is catching up. The U.K seems to be following the same path as the United States as well. On the other hand, in Poland the gap seems to be closing because the bottom SES group is catching up much faster. The Netherlands shows a similar pattern as the U.S, U.K patterns but upon closer inspection the explanation is very different. We see that the slope is negative (like in the previous two countries) but that's because the low SES group is decreasing at a greater rate than the high SES is decreasing.

% Make country summaries: Like the US is decreaing but because of bla bla (summing up everything up until now). Look at the graph and build rough summaries.

@@

We see that there are more countries where High SES is increasing inequality faster than low SES is catching up (top-left panel) but there's also a fair share where low SES is catching up faster, such as the United States and Germany. Regardless of this, we see an overall negative relationship where countries in which the low SES group is catching up, the high SES group goes down. This is very interesting because it suggests that there's a trend for countries to have the bottom SES groups are climbing upwards, a trait of meritocracy. However, this means that the top group comes down, rather than stagnate or increase. Equally important, we see very few countries where low SES are decreasing (so going even further down) and High SES is also decresing (bottom left panel). On the other hand, we do see more countries where the both groups are increasing (top right panel). We plot all remaining countries in the back, together with a trendline and we see that the relation holds for all PISA countries. To sum up, we see no definitive evidence of equality in all of these countries. Regardless of different tracking setups, we see that there's a tradeoff: either the bottom comes up or the top comes down or the top goes even further up and the bottom further down.

% Make country summary of each country for the tracking hypothesis #3. Like for example, countries in the bottom-right quadrant have low catching up and high coming down, do they have a common tracking structure? Do the same for the three quadrants.

@@

\subsection{Equality-efficiency tradeoff}

Moving on, we now discuss if there's a tradeoff between an educational system being efficient, i.e raising the overall performance of the country and whether that challenge compromises equality at the same time. For example, in order to improve the average score in, let's say, mathematics, tracking can improve the average score by making children in the higher trackers perform better, raising the overall performance. This, however, can come at a price. If tracking actually benefits the top but doesn't increase the performance of the bottom tracks then we see a net increase of inequality accompanied with an increase in the overall performance. Given that we have the data for the average performance for each country and the achievement gap, we can investigate this relationship in detail. We first being by plotting all countries pooled for each year in figure 8.

<<graphing_achievement_disparity, cache = T>>=

@@

For the past 15 years we see a very strong correlation suggesting that countries that perform very well in PISA have actually very low achievement gaps (or the other way around) \footnote{We graph all countries-years pooled as well and the relationship is identical}. The correlations are between \Sexpr{round(max(correlation), 2)} and \Sexpr{round(min(correlation), 2)}. This confirms the theoretical and empirical frameworks suggested by \citet{werfhorst_mijs}. However, what they didn't really go in depth was whether this corelation is strongest for some countries over time. However, we should not get trapped into an ecological fallacy and assume this relationship is identical within each country. With the data available we can look at the evolution separately for many countries. We plot it up next in figure 9.

@@

Looking at figure 9, we see a strong correspondence between the two variables. In countries like France we see quite a functional form between both inequality and performance. As average performance plummets, the achievement gap is increasing. This pattern is also seen in Austria, Australia and Germany. Yet we also see the opposite trend. We see countries like Poland, Finland, Canada and less pronounced, Italy, where this pattern is evident. The surprising case here is the United States. We see that as the average performance has been decreasing, the average achievement gap has been decreasing as well. In fact, among the few countries we should find something that resembles this pattern should be the U.S because it doesn't have a strict institutionalized tracking system. However, the explanation behind this pattern is very speculative.

@@

These results actually confirm that there exists a partial ecological fallacy. We see countries like Poland, Canada and Germany where the pattern is just as the aggregate trend, namely that as the average performance increases, inequality seems to be reduced in parallel. Conversely, France, Austria and Australia, to name a few, portray the opposite: as average performance decreases inequality increases. With these results we confirm our second hypothesis, that there is negative relationship. But we also unconver important variation within countries in the relationship between average achievement and dispersion of achievement.

\subsection{The role of curricular tracking in explaining inequalities}

@@

We now move to the final section of the paper, where we discuss what is a likely explanation to the patterns from before. The reason why countries differ in their evolution of inequality is still unknown. There is ample evidence showing the inequality between countries, to a certain extent, can be explained by the degree of tracking \citep{hanushek_woesmann_tracking}. But we are also interested in what explains the between country evolution of inequality, that is, the explanation as to why in certain countries it is increasing more than others over time.

This is clearly a difficult task given that, at least at the country-level, not many features of the educational system change in a 15-year time span, resulting in a practically invariant variable from which to explain the level of inequality (without any variance, we can't explain anything.). However, there are other strategies that can at least partially help us to understand the phenomena, and these are precisely the ones we follow.

Following the literature on the topic, we suspect that this evolution is in part very tightly related to the degree of stratification of the educational system. In a very simple frame-work, \citet{hanushek_woesmann_tracking} found that tracking indeed exarcebates inequality in the life course of children. Following the same strategy, we believe that tracking is also allowing the same gap in 15 year olds to grow over time, acompannied by other things. Yet, instead of concentrating only on the age of selection (just as they do), we explore a more fine grained definition of tracking, which is possible through the work of \citet{bol2013}. Aside from the age at first selection into tracks, we also use the the number of tracks in the country, the percentage of the entire curriculum that is tracked and a vocational index \footnote{This index is a factor loading from a principal factor analysis of the percentage of students in upper secondary vocational education (taken from two sources, to reduce measurement error). We take all of the data from \citet{bol2013}.}.

@@

In the final model, we regress the inequality of a country on a dummy variable where 1 equals only 1 track vs more than 1 track, whether the age of first selection is 15 or more vs below 15, a dummy stating whether the percentage of tracked curriculum is above 0 (that is, any tracking), and the standardized vocational index \footnote{We create dummies to avoid multicollinearity. The final model has a maximum VIF of 2.2.}. We are reluctant to include more variable in the model, first to keep it more parsimonious, and secondly to prevent overfitting due to our sample size. We present the results in the table below.

@@

Given that the classical p values are inexistent in bayesian modelling, we present uncertainty intervals next to each coefficient. The first model only includes the 1 track dummy. This variable shows that countries that have one track, have on average, half a standard deviation less inequality than other countries. This covariate alone explains about 23\% of the variance of the evolution of the 90/10 achievement gap. The second model now includes the dummy of whether the country has an age selection of 15 or more and we can see that countries with late selection into tracking have about -.38 less standard deviations in inequality. Also note that the number of tracks dummy was now reduced, suggesting these two variables are explaining similar things. This is expected given that late tracking is usually associated with fewer tracks, such as in the Scandinavian countries. The third model now includes whether the country has any degree of tracking, meaning whether the country has above 0\% of the curriculum tracked. The coding of the variable comes from visual inspection of the distribution, where some countries, such as the United States and Canada, have untracked curriculums. For this last variable we see no relationship at all, but after including the vocational index, the model is completely reversed with countries with tracked curriculum having about 0.6 more standard deviations in inequality that countries with no tracking.

We also note that vocational enrollment is associated with less inequality. Not surprisingly, the age of selection coefficient, also increased due to the correlation between age of selection and vocational enrollment. Countries with high vocational enrollment are also those with early age selection, such as Germany and Austria. These two variables should be inspected further, perhaps with an interaction, given that both coefficients are sizable and very correlated. This final model explains over one third of the variance in the evolution of the 90/10 SES gap, with an R square of 40\%. The uncertainty of each of the covariates varies a lot with the percentage of curriculum tracked dummy ranging all the way from 0.9 to 0.3. Nonetheless, we see that the other coefficients are very strong and have reasonable intervals. It is very important to consider the uncertainty of our estimates, to gain confidence in our model fit. So far, this model seems to fit the data quite reasonably.

All of these countries have a different tracking structure, with some setups being more egalitarian than others. Using the previous model we make a simulation and predict the level of the 90/10 achievement gap for all possible tracking setups for all countries. Below we plot all these combinations. The X axis has the number of tracks that a country has, the Y axis has the predicted 90/10 achievement gap and below each country name we have whether the country has an age selection of 15 or more or below 15. In addition, to confirm how accurate the model is, we also plot the actual level of inequality in the actual tracking setup of the country. Let's take Germany for example.

@@

The red point for Germany is in the '\textgreater 1 track' and '\textless 15 age' setup, meaning that Germany has a tracking system that has more than 1 track and an age of selection of below 15. We see that the model predicts the level of inequality for that set up quite accurately, relative to the actual level (red point). But if the setup would be in the ideal '1 track' and age of selection above or equal to 15, the inequality would be much lower. Generally speaking, this applies to all countries in the plot. Virtually for all countries which are not in this 'ideal' tracking, a switch to the 'ideal' would reflect a reduction of the achievement gap.

% Comment on the above graph for particular countries! Make some observations

For those countries which are not in the 'ideal' tracking setup, we compute below the percentage change from switching to the ideal tracking.

@@

The biggest reductions are for both Hungary and Germany, with 46\% and 36\% respectively. France, which had a particularly steep increase in the last 15 years, would see a reduction of the 90/10 SES gap of 31\% and Austria, with a very similar tracking structure as the German case, would see their achievement gap reduced by 20\%. The very suprising result is the Dutch case, for which a switch from the 'worst' tracking setup to the 'best' case would result in an increase of the gap of 19\%.

This last result could mean two things. First and foremost, the Dutch educational system has built other mechanisms that, despite the tracking system, help children from low social origins escalate upwards. This explanation is not very trivial considering that in most mobility research, the Netherlands stands out as a very equal country relative to other countries with a similary tracking structure, like Germany \citep{shavit1993, esping2012}. Another possible explanation, and something which needs to assessed, is whether our model is predicting accurately. If we look at the plot from above, we see that for some countries, the model is a bit off from the actual level of inequality. Moreover, although we see that if all countries switched to the 'ideal' tracking setup we would see a vast reduction in inequality, we also see that each of these predictions is followed by a wide uncertainty interval.

Considering that the results are very consistent for every country, and in most countries the prediction matches the actual value, this could simply be a consequence of low sample size. Whether these results stand the test of a different empirical strategy or a higher sample size, we leave up to future research.

A more summarized empirical strategy is to actually standardize all the tracking indicator into one index, something that \citet{bol2013} already do quite nicely. Using this standardized tracking variable and the same bayesian multilevel model, we now model the tracking and vocational index as before, but add an interaction between two, given that in the previous model we recorded a strong shift in the model when tracking-related variables and the vocational model where both included in the same model. Below are the results for this model.

@@

We confirm the results from the previous models, where more tracking is associated with an achievement gap of about 0.4 standard deviations wider and more vocational enrollment is associated with a reduction of about 0.2 standard deviations, half the slope of tracking. Meaning that the tracking effect might be offsetting the effect of the vocational index given that it's half the size of the slope. Both these coefficients are actually significantly different from each other, so it is possible. Before we explore it visually, note that the interaction between the two is very reliable and significant, considering that small sample size. Below we graph the interaction for different quantiles of the vocational index.
